Functional traits complicate the picture of temporal biodiversity change in bird and mammal communities

### 2024-03-26 paper/GEB Submission/manuscript_GEB.tex
Add table, change text color

\textcolor{blue}{Using general trends derived from limited data as a diagnostic for the state of biodiversity or directing conservation is a topic of ongoing debate.}
Global analyses have been heavily criticized for geographic biases, lack
of data in the most heavily impacted areas, and exclusion of individual
studies' ecological context (10--12). Many of these criticisms reflect
limitations of ecological data on the whole,
\textcolor{blue}{leading to a call for additional data not only to fill geographic and temporal gaps, but to flesh out key characteristics of communities}
(13, 14).

@@

\textcolor{blue}{The expectation for functional change across communities is not obvious from past work, and may or may not follow taxonomic trends}
(13). While \textcolor{blue}{loss of functional diversity} is frequently
cited as one of the most pressing concerns of the anthropocene (18--20),
\textcolor{blue}{functional diversity may be maintained even when species are lost from a community}
(21, 22). Forecasts of functional loss range from negligible (23) to
dire (24, 25). And while some observed trends show significant
functional loss (26) others document no loss even in some of the most
heavily impacted communities (27--29). On paleoecological time scales,
\textcolor{blue}{the functional space} shows mixed responses to
environmental change and extinction events (30, 31), with significant
impacts of species extinctions on functional diversity in some taxa and
not others (32). For some time periods, functional diversity appears to
be maintained for substantial portions of geological time (33).
\textcolor{blue}{Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds [@barnagaud2017; @jarzyna2016] and ray-finned fishes, sharks, and rays}
(34).
\textcolor{blue}{Additionally, these taxa have been the subject of substantial conservation investment and policy action, so are not necessarily representative of trends for other taxa or the same taxa in different locations.}

Here we
\textcolor{blue}{leverage ongoing efforts to assemble functional trait data and recent computational advances to}
perform the first multi-taxa, multi-realm assessment of functional
diversity change through time. We focus on mammal and bird species as
\textcolor{blue}{a subset of the world's biodiversity of particular conservation concern that is heavily impacted by anthropogenic change.}While
examining trends in plants, invertebrates, and other vertebrate species
is of equal interest, trait data for those taxa raise additional
challenges such as limited and biased species coverage (35), a lack of
accepted species-level means, and differences in the types of traits
collected. To ensure comparability across taxa in trait type and data
quality we therefore focus on mammals and birds.
\textcolor{blue}{We include body mass, foraging and dietary traits} that
were intentionally selected to be representative of a species' Eltonian
niche, thereby summarizing the functional role they play in the
community (36). An initial assessment of amphibian trends is included in
the supplement, but excluded from general trend assessment here due to
limited geographic coverage.

We assess thousands of mammal and bird functional diversity time series
to determine whether or not
\textcolor{blue}{the addition of functional trait data gives a clarifying picture of biodiversity change across communities. Rather than testing specific hypotheses of change we present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination. We present results at three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.}

@@

We obtained mammal and bird time series from the BioTIME database, a
global repository of high-quality assemblage time series
\textcolor{blue}{collected from the literature and ongoing monitoring efforts}.
\textcolor{blue}{Data is structured such that a study encompass all data collected following consistent sampling protocols and may contain multiple time series from separate locations. }
Time series represent full assemblages rather than populations of single
species (31), and all time series include abundance of observed species.
Following best practices for the database (37), studies with multiple
sample locations were split into individual time series following a
standardized spatial scale. Scale was set by a global grid with cell
size determined based on the sample extent of studies with only a single
location (see 31 for details on how sample extents were defined), with
the area of each cell set to one standard deviation away from the mean
of the single extent locations. All samples from a study within a single
cell were considered to be a single time series, and species abundances
were combined for all samples.

We used trait data from the Elton Trait Database, which consists of
species-level means for traits that represent species' multifaceted role
in the community (36). Traits include: body mass, diet, active diel
period, nocturnality, forest foraging strata, pelagic use. Multiple
traits (i.e.~diet, foraging strata, activity seasonality, active diel
period) were broken down into percentage or binary use for each level.

In order to ensure taxonomic consistency across datasets, BioTIME
species were paired with trait data based on their species identifier
from the Integrated Taxonomic Information System database (retrieved
09-15-2020 from the on-line database,
\url{https://doi.org/10.5066/F7KH0KBK}), obtained through the
\texttt{taxadb} R package (38, 39). If more than one species in the
assemblage data resolved to the same identifier, observations were
considered the same species. For trait data, traits for all species of
the same identifier were averaged. Only studies for which at least 75\%
of species had trait data were included. In order to have a sufficient
number of species to calculate functional diversity metrics, years with
fewer than 5 species observed were also excluded. Sensitivity analyses
were conducted for the trait coverage threshold and the duration of
included time series.

@@

Functional diversity metrics were calculated using the \emph{dbFD}
function from the \emph{FD} R package (40). Here we report functional
richness (\emph{FRic}), functional evenness (\emph{FEve}), and
functional divergence (\emph{FDiv}) which together describe three
complementary characteristics of the functional space (41, 42).
\emph{FRic} assesses the volume of the trait space occupied by species
in the community, with higher values indicating communities with species
of more extreme trait values. \emph{FEve} describes how species are
distributed across the trait space and how abundance is distributed
across species. Higher values of \emph{FEve} indicate more even spacing
of species in the trait space and individuals across species.
\emph{FDiv} measures the degree to which species and their abundances
maximize differences in the functional space. Higher values of
\emph{FDiv} therefore correspond to communities where many highly
abundant species are on the edges of the trait space. We also calculated
the community-weighted mean (\emph{CWM}) of included traits to examine
shifts in the distribution of each trait.
\textcolor{blue}{Hereafter we refer to results for functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).}

All available trait data for each study were included in functional
diversity calculations with the exception of traits that were the same
value for all observed species in the study. For variables with multiple
levels each level was included as a separate trait axis. Continuous
traits were z-score scaled to give each trait equal weight in the trait
space (43, 44).
\textcolor{blue}{Before calculating diversity metrics, dbFD reduces the dimensionality of the trait space by performing PCoA. We limited the number of included PCoA axes to}
the maximum number of traits that fulfills the criteria \(s >= 2^t\),
where \(s\) is the number of species and \emph{t} is the number of
traits. This restriction allows for enough axes to capture the trait
space while maintaining computational feasibility (45). Metrics
incorporated weighting based on species abundance.

@@

To assess functional change independent of species richness we
calculated the standardized effect size (SES) for each of the three
functional diversity metrics (\emph{FRic}, \emph{FEve}, \emph{FDiv})
from null estimates (46). Null model corrections allow us to assess the
degree to which the observed functional diversity metric deviates from
the value expected by chance in a randomly assembled community. Null
estimates were calculated for each rarefied sample by randomly sampling
species from the species pool for each year and randomly assigning
observed abundances to species. `
\textcolor{blue}{Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability}.
This process was repeated 500 times to get an estimate and standard
deviation of the null expectation for the metric for each rarefaction
sample for that time series. We used these values to calculate SES using
the following formula:
\(SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}\). We then
calculated the median SES estimate for each metric from all the
rarefaction samples for a time series. SES estimates can be interpreted
as how much of the functional characteristic (richness, evenness,
divergence) was observed beyond what was expected by chance for a
community of that species richness.
\textcolor{blue}{This approach will be less accurate for shorter time series, as we likley will not have captured all available species in the true species pool, but it is impossible to know whether the mean estimate from the null model is an over or under estimate without knowing the functional characteristics of the missing species.}

@@

We estimated general trends
\textcolor{blue}{across bird and mammals communities} for each diversity
metric using a linear mixed effects model with a random slope and
intercept for each study and each time series nested within the study,
\textcolor{blue}{methods which deal well with the inherent imbalances in our data.}
We fit 18 individual \emph{CWM} models,
\textcolor{blue}{one for each trait included in the analsysis}. All time
series with data for a given trait were included in the corresponding
\emph{CWM} model. We estimated study level trends using individual
linear models. For studies with more than one times series we fit a
random slope and intercept for time series. Some study-level models
could not be fit for five studies for at least one metric due to data
limitations, but those studies were still included in the general
models. They represented 12 of 1350 study-level models fit for each
metric. For further details see the supplement. Where appropriate,
response variables were \(log\) or \(log(x+1)\) transformed to better
fit model assumptions of residual normality.

To test for trends within and between different levels of taxa, biome,
realm, and protection status we fit separate models with each of those
factors added as a predictor interacting with time to the original model
structure. We estimated within-level slopes and calculated between-level
contrasts using the \emph{emmeans} package (v1.8.2, 47).
\textcolor{blue}{For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore refrained from interpreting results for levels where there were fewer than three studies.}
We assessed the impact of time series duration and start year on
study-level trends using linear models with duration and start year as
predictors. All models in our analysis were fit using the \emph{lme4}
(v1.1-30) package in R (v4.2.3) and p-values were calculated by
Satterthwaite's degrees of freedom method using the \emph{lmerTest}
(v3.1-3) package with a significance level of \(\alpha = 0.05\)
(48--50).

@@

We found no significant overall \textcolor{blue}{temporal} trend in
species richness or functional diversity metrics
\textcolor{blue}{including functional richness, evenness, or divergence}
(observed or \textcolor{blue}{corrected}) (Fig
\ref{fig:timeseriesPlot}). We did find a significant overall decrease in
Jaccard similarity, indicating accumulating changes in species
composition. Non-significant overall \textcolor{blue}{temporal}trends
indicate that although some studies experience increasing or decreasing
trends, the average trend across studies
\textcolor{blue}{was not significantly different from zero}` (Table
\ref{tab:resultsTab}). \textcolor{blue}{Within-group} trends for
different taxa, biomes, realms, or protection statuses were also
non-significant for richness and functional diversity metrics, with the
exception of a significantly increasing trend for functional evenness of
marine studies, and a significantly decreasing functional richness slope
for mammal studies. However, trends were not exhibited for the
\textcolor{blue}{corrected} metric indicating that differences
\textcolor{blue}{in functional diversity metrics} were largely due to
changes in species richness. The general trends for \emph{CWM} models
were similarly not significant, with a significant positive trend for
only percentage of fish in diet composition.

@@

@@

@@

\textcolor{blue}{The most stark area of consensus was the lack of general trend aggregating across communities.}
We did not detect an overall trend in any functional diversity metric,
\textcolor{blue}{corrected or uncorrected}. As with previous
species-based syntheses, we also found no overall trend in species
richness accompanied by \textcolor{blue}{increasing dissimilarity}
through time (31), indicating that non-significant trends in functional
metrics are consistent with similar well-documented species derived
trends. We found significant turnover for all biomes, realm, and taxa
except for marine studies, which stands in contrast to other global
estimates of biodiversity change that found higher turnover in marine
systems than terrestrial (37). However, previous global estimates are
dominated by fish communities which we exclude here and may be driving
the overall trend while disguising relative stasis in marine bird and
mammal communities.

We found evidence of functional richness loss in mammal studies, with a
loss trend significantly different from zero and significantly different
from the trend for birds, indicating a potential loss of functional
capacity in these communities. This result is consistent with previous
work linking anthropogenic drivers such as habitat loss, poaching, and
human modification and development to loss of functional diversity in
mammal communities across the globe (51--53). However we did not find
trends outside the random expectation for species loss, in contrast to
expectations for multiple continents based on IUCN risk categories (51).
In general, functional capacity of mammals shows greater declines that
bird communities.

We also found evidence of increasing evenness for Marine studies. The
empirical link between changes in evenness and ecosystem process is the
most poorly studied of the metrics, however theory indicates that
increased evenness improves function and stability (54). If one or a few
species become largely dominant in a community, function will be the
result of only those species rather than spread out across a wealth of
species responding independently to environmental change (42, 55).
Further, there is some evidence that maintaining evenness is important
for supporting multifunctionality in communities (56). The majority of
marine studies are of seabird communities (7 of 9), so the general trend
may be an early indicator of pay off from investments in sea bird
recovery, which is also indicated in recent species-based work (57).

Compositional shifts shown by significant trends in \emph{CWM'}s were
restricted to dietary traits, with no evidence of a trend in foraging
traits or body mass. Increases in fish consumption for bird and
temperate studies my be the product of increased availability of easy
fish from commercial fishing operations as fish are processed or
stocking of inland waterways for sport fishing. However, fisheries also
have negative impacts on waterbirds through competition for fish or bird
mortality due to by catch. Decreases in nectar consumption for tropical
studies is likely due declines in bat species, as 4 of the 5 tropical
studies are for mammal and bats were the only nectar consuming mammals
sampled in our data. Interpretation of the increases for temperate
studies is less clear, as temperate studies included many bird
communities with a variety of nectar consumers. Similarly, decreasing
trends for vertebrate consumption in marine and tropical studies are not
obviously consistent with known changes in those contexts and warrant
further examination.

@@

While we found no over all trends in functional metrics, our results
should not be interpreted as an indication that the ongoing biodiversity
crisis is less severe than previously described, or that there is no
concern for functional change as a result of anthropogenic impact. In
fact, study-level trends indicate quite the opposite, that functional
shifts with negative or yet unknown implications for ecosystem processes
may be going undetected by common species-based approaches. For example,
loss of evenness in communities with constant species richness may be an
first sign of a community being impacted by environmental change, with
negative implications for stability and function.

>>>>

\textcolor{blue}{Using general trends derived from limited data as a diagnostic for the state of biodiversity or directing conservation action is a topic of ongoing debate.}
Global analyses have been heavily criticized for geographic biases, lack
of data in the most heavily impacted areas, and exclusion of individual
studies' ecological context (10--12). Many of these criticisms reflect
limitations of ecological data on the whole,
\textcolor{blue}{leading to a call for additional data not only to fill geographic and temporal gaps, but to flesh out key characteristics of communities}
(13, 14).

@@

\textcolor{blue}{The expectation for functional change across communities is not obvious from past work, and may or may not follow taxonomic trends}
(13). While \textcolor{blue}{loss of functional diversity} is frequently
cited as one of the most pressing concerns of the anthropocene (18--20),
\textcolor{blue}{functional diversity may be maintained even when species are lost from a community}
(21, 22). Forecasts of functional loss range from negligible (23) to
dire (24, 25). And while some observed trends show significant
functional loss (26) others document no loss even in some of the most
heavily impacted communities (27--29). On paleoecological time scales,
\textcolor{blue}{the functional space} shows mixed responses to
environmental change and extinction events (30, 31), with significant
impacts of species extinctions on functional diversity in some taxa and
not others (32). For some time periods, functional diversity appears to
be maintained for substantial portions of geological time (33).
\textcolor{blue}{Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds}
(34, 35) \textcolor{blue}{and ray-finned fishes, sharks, and rays} (36).

Here we
\textcolor{blue}{leverage ongoing efforts to assemble functional trait data and recent computational advances to}
perform the first multi-taxa, multi-realm assessment of functional
diversity change through time. We focus on mammal and bird species as
\textcolor{blue}{a subset of the world's biodiversity of particular conservation concern that is heavily impacted by anthropogenic change.}While
examining trends in plants, invertebrates, and other vertebrate species
is of equal interest, trait data for those taxa raise additional
challenges such as limited and biased species coverage (37), a lack of
accepted species-level means, and differences in the types of traits
collected. To ensure comparability across taxa in trait type and data
quality we therefore focus on mammals and birds.
\textcolor{blue}{We include body mass, dietary, foraging and other behavioral traits}
that were intentionally selected to be representative of a species'
Eltonian niche, thereby summarizing the functional role they play in the
community (38). An initial assessment of amphibian trends is included in
the supplement, but excluded from general trend assessment here due to
limited geographic coverage.

We assess thousands of mammal and bird functional diversity time series
to determine whether or not
\textcolor{blue}{the addition of functional trait data gives a clarifying picture of biodiversity change across communities. Rather than testing specific hypotheses of change we present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination. We present results for three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.}

@@

We obtained mammal and bird time series from the BioTIME database, a
global repository of high-quality assemblage time series
\textcolor{blue}{collected from the literature and ongoing monitoring efforts}.
\textcolor{blue}{Data is structured such that a study encompass all data collected following consistent sampling protocols and may contain multiple time series from separate locations.}
Time series represent full assemblages rather than populations of single
species (31), and all time series include abundance of observed species.
Following best practices for the database (39), studies with multiple
sample locations were split into individual time series following a
standardized spatial scale. Scale was set by a global grid with cell
size determined based on the sample extent of studies with only a single
location (see 31 for details on how sample extents were defined), with
the area of each cell set to one standard deviation away from the mean
of the single extent locations.
\textcolor{blue}{The resulting cell size for our data was approximately 95 }km\textsuperscript{2}.
All samples from a study within a single cell were considered to be a
single time series, and species abundances were combined for all
samples.

We used trait data from the Elton Trait Database, which consists of
species-level means for traits that represent species' multifaceted role
in the community (38). Traits include: body mass, diet, nocturnality,
forest foraging strata, pelagic use. With the exception of body mass,
traits were broken down into percentage or binary use for each level of
the trait type (Table \ref{tab:traitTable}).

\begin{table}

\caption{\label{tab:traitTable}Description of the traits included in the analysis broken down by categories at data type.}
\centering
\begin{tabular}[t]{l|l|l}
\hline
Trait & Category & Data Type\\
\hline
 & Invertebrate & \\
\cline{2-2}
 & Mammals and Birds & \\
\cline{2-2}
 & Reptiles & \\
\cline{2-2}
 & Fish & \\
\cline{2-2}
 & Unknown Vertebrates & \\
\cline{2-2}
 & Scavenging & \\
\cline{2-2}
 & Fruit & \\
\cline{2-2}
 & Nectar & \\
\cline{2-2}
 & Seeds & \\
\cline{2-2}
\multirow{-10}{*}{\raggedright\arraybackslash Diet} & Other Plant & \multirow{-10}{*}{\raggedright\arraybackslash percentage consumed}\\
\cline{1-3}
 & Below water surface & \\
\cline{2-2}
 & water surface & \\
\cline{2-2}
 & ground & \\
\cline{2-2}
 & understory & \\
\cline{2-2}
 & > 2m, below canopy & \\
\cline{2-2}
 & canopy & \\
\cline{2-2}
\multirow{-7}{*}{\raggedright\arraybackslash Foraging Strata} & aerial & \multirow{-7}{*}{\raggedright\arraybackslash percentage of use}\\
\cline{1-3}
 & yes & \\
\cline{2-2}
\multirow{-2}{*}{\raggedright\arraybackslash Pelagic Specialist} & no & \\
\cline{1-2}
 & yes & \\
\cline{2-2}
\multirow{-2}{*}{\raggedright\arraybackslash Nocturnal} & no & \multirow{-4}{*}{\raggedright\arraybackslash binary}\\
\cline{1-3}
Body Mass & - & continous, in grams\\
\hline
\end{tabular}
\end{table}

In order to ensure taxonomic consistency across datasets, BioTIME
species were paired with trait data based on their species identifier
from the Integrated Taxonomic Information System database (retrieved
09-15-2020 from the on-line database,
\url{https://doi.org/10.5066/F7KH0KBK}), obtained through the
\texttt{taxadb} R package (40, 41). If more than one species in the
assemblage data resolved to the same identifier, observations were
considered the same species. For trait data, traits for all species of
the same identifier were averaged. Only studies for which at least 75\%
of species had trait data were included. In order to have a sufficient
number of species to calculate functional diversity metrics, years with
fewer than 5 species observed were also excluded. Sensitivity analyses
were conducted for the trait coverage threshold and the duration of
included time series.

@@

Functional diversity metrics were calculated using the \emph{dbFD}
function from the \emph{FD} R package (42). Here we report functional
richness (\emph{FRic}), functional evenness (\emph{FEve}), and
functional divergence (\emph{FDiv}) which together describe three
complementary characteristics of the functional space (43, 44).
\emph{FRic} assesses the volume of the trait space occupied by species
in the community, with higher values indicating communities with species
of more extreme trait values. \emph{FEve} describes how species are
distributed across the trait space and how abundance is distributed
across species. Higher values of \emph{FEve} indicate more even spacing
of species in the trait space and individuals across species.
\emph{FDiv} measures the degree to which species and their abundances
maximize differences in the functional space. Higher values of
\emph{FDiv} therefore correspond to communities where many highly
abundant species are on the edges of the trait space. We also calculated
the community-weighted mean (\emph{CWM}) of included traits to examine
shifts in the distribution of each trait.
\textcolor{blue}{Hereafter we refer to results for functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).}

All available trait data for each study were included in functional
diversity calculations with the exception of traits that were the same
value for all observed species in the study. For variables with multiple
levels each level was included as a separate trait axis. Continuous
traits were z-score scaled to give each trait equal weight in the trait
space (45, 46).
\textcolor{blue}{Before calculating diversity metrics, dbFD reduces the dimensionality of the trait space by performing PCoA. We limited the number of included PCoA axes to}
the maximum number of traits that fulfills the criteria \(s >= 2^t\),
where \(s\) is the number of species and \emph{t} is the number of
traits. This restriction allows for enough axes to capture the trait
space while maintaining computational feasibility (47). Metrics
incorporated weighting based on species abundance.

@@

To assess functional change independent of species richness we
calculated the standardized effect size (SES) for each of the three
functional diversity metrics (\emph{FRic}, \emph{FEve}, \emph{FDiv})
from null estimates (48). Null model corrections allow us to assess the
degree to which the observed functional diversity metric deviates from
the value expected by chance in a randomly assembled community. Null
estimates were calculated for each rarefied sample by randomly sampling
species from the species pool for each year and randomly assigning
observed abundances to species.
\textcolor{blue}{Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability}.
This process was repeated 500 times to get an estimate and standard
deviation of the null expectation for the metric for each rarefaction
sample for that time series. We used these values to calculate SES using
the following formula:
\(SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}\). We then
calculated the median SES estimate for each metric from all the
rarefaction samples for a time series. SES estimates can be interpreted
as how much of the functional characteristic (richness, evenness,
divergence) was observed beyond what was expected by chance for a
community of that species richness.
\textcolor{blue}{This approach will be less accurate for shorter time series, as we likley will not have captured all available species in the true species pool, but it is impossible to know whether the mean estimate from the null model is an over or under estimate without knowing the functional characteristics of the missing species.}

@@

We estimated general trends
\textcolor{blue}{across bird and mammals communities} for each diversity
metric using a linear mixed effects model with a random slope and
intercept for each study and each time series nested within the study,
\textcolor{blue}{methods which deal well with the inherent imbalances in our data.}
We fit 18 individual \emph{CWM} models,
\textcolor{blue}{one for each trait included in the analsysis}. All time
series with data for a given trait were included in the corresponding
\emph{CWM} model. We estimated study level trends using individual
linear models. For studies with more than one times series we fit a
random slope and intercept for time series. Some study-level models
could not be fit for five studies for at least one metric due to data
limitations, but those studies were still included in the general
models. They represented 13 of 1350 study-level models fit for each
metric. For further details see the supplement. Where appropriate,
response variables were \(log\) or \(log(x+1)\) transformed to better
fit model assumptions of residual normality.

To test for trends within and between different levels of taxa, biome,
realm, and protection status we fit separate models with each of those
factors added as a predictor interacting with time to the original model
structure. We estimated within-level slopes and calculated between-level
contrasts using the \emph{emmeans} package (v1.8.2, 49).
\textcolor{blue}{For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore refrained from interpreting results for levels where there were fewer than three studies.}
We assessed the impact of time series duration and start year on
study-level trends using linear models with duration and start year as
predictors. All models in our analysis were fit using the \emph{lme4}
(v1.1-30) package in R (v4.2.3) and p-values were calculated by
Satterthwaite's degrees of freedom method using the \emph{lmerTest}
(v3.1-3) package with a significance level of \(\alpha = 0.05\)
(50--52).

@@

We found no significant overall \textcolor{blue}{temporal} trend in
species richness or functional diversity metrics
\textcolor{blue}{including functional richness, evenness, or divergence}
(observed or \textcolor{blue}{corrected}) (Fig
\ref{fig:timeseriesPlot}). We did find a significant overall decrease in
Jaccard similarity, indicating accumulating changes in species
composition. Non-significant overall \textcolor{blue}{temporal} trends
indicate that although some studies experience increasing or decreasing
trends, the average trend across studies
\textcolor{blue}{was not significantly different from zero}` (Table
\ref{tab:resultsTab}). \textcolor{blue}{Within-group} trends for
different taxa, biomes, realms, or protection statuses were also
non-significant for richness and functional diversity metrics, with the
exception of a significantly increasing trend for functional evenness of
marine studies, and a significantly decreasing functional richness slope
for mammal studies. However, trends were not exhibited for the
\textcolor{blue}{corrected} metric indicating that differences
\textcolor{blue}{in functional diversity metrics} were largely due to
changes in species richness. The general trends for \emph{CWM} models
were similarly not significant, with a significant positive trend for
only percentage of fish in diet composition.

@@

\begin{table}[!h]

\caption{\label{tab:resultsTab}Model estimates and statistics for general trend models for species richness, Jaccard similarity, and corrected functional diversity metrics. Additional model estimates, including CWM models, can be found in the supplement.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{l|l|>{\raggedright\arraybackslash}p{2cm}|l|r|r|l}
\hline
metric & effect & grouping & term & estimate & std.error & p.value\\
\hline
 &  &  & Intercept & -0.25 & 0.11 & 0.03\\
\cline{4-7}
 & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & 0.00 & 0.04 & 0.96\\
\cline{2-7}
 &  &  & SD Intercept & 0.60 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.11 &  & \\
\cline{4-7}
 &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & -0.16 &  & \\
\cline{3-7}
 &  &  & SD Intercept & 0.60 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.23 &  & \\
\cline{4-7}
\multirow{-8}{*}{\raggedright\arraybackslash SES\_FDiv} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & -0.01 &  & \\
\cline{1-7}
 &  &  & Intercept & 0.10 & 0.17 & 0.56\\
\cline{4-7}
 & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & -0.01 & 0.02 & 0.62\\
\cline{2-7}
 &  &  & SD Intercept & 1.08 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.04 &  & \\
\cline{4-7}
 &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & -0.53 &  & \\
\cline{3-7}
 &  &  & SD Intercept & 0.40 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.17 &  & \\
\cline{4-7}
\multirow{-8}{*}{\raggedright\arraybackslash SES\_FEve} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & -0.21 &  & \\
\cline{1-7}
 &  &  & Intercept & -0.26 & 0.07 & <0.001\\
\cline{4-7}
 & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & 0.04 & 0.04 & 0.35\\
\cline{2-7}
 &  &  & SD Intercept & 0.29 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.11 &  & \\
\cline{4-7}
 &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & -0.27 &  & \\
\cline{3-7}
 &  &  & SD Intercept & 0.54 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.18 &  & \\
\cline{4-7}
\multirow{-8}{*}{\raggedright\arraybackslash SES\_FRic} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & 0.06 &  & \\
\cline{1-7}
 &  &  & Intercept & 0.46 & 0.02 & <0.001\\
\cline{4-7}
 & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & -0.03 & 0.00 & <0.001\\
\cline{2-7}
 &  &  & SD Intercept & 0.09 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.02 &  & \\
\cline{4-7}
 &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & -0.36 &  & \\
\cline{3-7}
 &  &  & SD Intercept & 0.08 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.02 &  & \\
\cline{4-7}
\multirow{-8}{*}{\raggedright\arraybackslash log(Jaccard + 1)} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & 0.05 &  & \\
\cline{1-7}
 &  &  & Intercept & 2.40 & 0.10 & <0.001\\
\cline{4-7}
 & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & -0.06 & 0.05 & 0.17\\
\cline{2-7}
 &  &  & SD Intercept & 0.64 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.27 &  & \\
\cline{4-7}
 &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & -0.70 &  & \\
\cline{3-7}
 &  &  & SD Intercept & 0.21 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.09 &  & \\
\cline{4-7}
\multirow{-8}{*}{\raggedright\arraybackslash log(Species Richness)} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & 0.34 &  & \\
\hline
\end{tabular}}
\end{table}

@@

\begin{table}

\caption{\label{tab:trendTab}Number of studies that experienced a significant trend in each calculated metric out of 50 total studies.}
\centering
\begin{tabular}[t]{ccccccccc}
\toprule
 & S & Jaccard Similarity & FRic & FEve & FDiv & SES FRic & SES FEve & SES FDiv\\
\midrule
+ & 7 & 0 & 7 & 5 & 3 & 4 & 4 & 3\\
- & 9 & 11 & 8 & 6 & 5 & 2 & 6 & 2\\
\bottomrule
\end{tabular}
\end{table}

@@

\textcolor{blue}{The most stark area of consensus was the lack of general trend aggregating across communities. We did not detect an overall trend in any functional diversity metric, corrected or uncorrected. As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing dissimilarity through time}
(31)\textcolor{blue}{, indicating that non-significant trends in functional metrics are consistent with similar well-documented species derived trends. We found significant turnover for all biomes, realm, and taxa except for marine studies, which stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial}
(39).
\textcolor{blue}{However, previous global estimates are dominated by fish communities which we exclude here and may be driving the overall trend while disguising relative stasis in marine bird and mammal communities.}

\textcolor{blue}{We found evidence of functional richness loss in mammal studies, with a loss trend significantly different from zero and significantly different from the trend for birds, indicating a potential loss of functional capacity in these communities. This result is consistent with previous work linking anthropogenic drivers such as habitat loss, poaching, and human modification and development to loss of functional diversity in mammal communities across the globe}
(53--55).
\textcolor{blue}{However we did not find trends outside the random expectation for species loss, in contrast to expectations for multiple continents based on IUCN risk categories}
(53).
\textcolor{blue}{In general, functional capacity of mammals shows greater declines that bird communities.}

\textcolor{blue}{We also found evidence of increasing evenness for Marine studies. The empirical link between changes in evenness and ecosystem process is the most poorly studied of the metrics, however theory indicates that increased evenness improves function and stability }
(56),
\textcolor{blue}{as the dominance of one or a few species makes function sensitive to the ability of those species to respond to environmental change }(44,
57).
\textcolor{blue}{Further, there is some evidence that maintaining evenness is important for supporting multifunctionality in communities }(58).
\textcolor{blue}{The majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery, which is also indicated in recent species-based work }(59).

\textcolor{blue}{Compositional shifts shown by significant trends in CWM's were restricted to dietary traits, with no evidence of a trend in foraging traits or body mass. Increases in fish consumption for bird and temperate studies my be the product of increased availability of easy fish from commercial fishing operations as fish are processed or stocking of inland waterways for sport fishing. However, fisheries also have negative impacts on waterbirds through competition for fish or bird mortality due to by catch. Decreases in nectar consumption for tropical studies is likely due declines in bat species, as 4 of the 5 tropical studies are for mammal and bats were the only nectar consuming mammals sampled in our data. Interpretation of the increases for temperate studies is less clear, as temperate studies included many bird communities with a variety of nectar consumers. Similarly, decreasing trends for vertebrate consumption in marine and tropical studies are not obviously consistent with known changes in those contexts and warrant further examination.}

@@

\textcolor{blue}{While we found no over all trends in functional metrics, our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact. In fact, study-level trends indicate quite the opposite, that functional shifts with negative or yet unknown implications for ecosystem processes may be going undetected by common species-based approaches. For example, loss of evenness in communities with constant species richness may be an first sign of a community being impacted by environmental change, with negative implications for stability and function.}

\textcolor{blue}{One of the biggest threats to biodiversity is the wholesale conversion of natural areas to urban or human-dominated landscapes }(77).
\textcolor{blue}{Typical long term monitoring such as those included in our study stops before this conversion occurs, leaving the resultant precipitous declines in biodiversity unrecorded}
(10).
\textcolor{blue}{This is a known issue with the culture of long term monitoring, and our results should not be removed from that context. Rather, this studies captures communities that are likely experiencing a degree of human intervention, but are still largely nature-dominated.}
